Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions

### 2022-11-22 StatPaper/APA-lato_New/APA/wileyNJD-APA.tex
made a version, need to readover

\title{This is the sample article title\protect\thanks{This is an example for title footnote.}}

\author[1]{Author One*}

\author[2,3]{Author Two}

\author[3]{Author Three}

\authormark{AUTHOR ONE \textsc{et al}}

\address[1]{\orgdiv{Org Division}, \orgname{Org Name}, \orgaddress{\state{State name}, \country{Country name}}}

\address[2]{\orgdiv{Org Division}, \orgname{Org Name}, \orgaddress{\state{State name}, \country{Country name}}}

\address[3]{\orgdiv{Org Division}, \orgname{Org Name}, \orgaddress{\state{State name}, \country{Country name}}}

\corres{*Corresponding author name, This is sample corresponding address. \email{authorone@gmail.com}}

@@

\abstract[Summary]{This is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text this is sample abstract text.}

\keywords{keyword1, keyword2, keyword3, keyword4}

\jnlcitation{\cname{%
\author{Williams K.},
\author{B. Hoskins},
\author{R. Lee},
\author{G. Masato}, and
\author{T. Woollings}} (\cyear{2016}),
\ctitle{A regime analysis of Atlantic winter jet variability applied to evaluate HadGEM3-GC2}, \cjournal{Q.J.R. Meteorol. Soc.}, \cvol{2017;00:1--6}.}

@@

\footnotetext{\textbf{Abbreviations:} ANA, anti-nuclear antibodies; APC, antigen-presenting cells; IRF, interferon regulatory factor}

\section{Sample for first level head}\label{sec1}

Lorem ipsum dolor sit amet, consectetuer adipiscing elit~\citet{Rothermel1997} Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae,
felis. Curabitur dictum gravida mauris. Nam arcu libero, nonummy eget, consectetuer id, vulputate a, magna. Donec
vehicula augue eu neque. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.
Mauris ut leo. Cras viverra metus rhoncus sem. Nulla et lectus vestibulum urna fringilla ultrices. Phasellus eu tellus
sit amet tortor gravida placerat. Integer sapien est, iaculis in, pretium quis, viverra ac, nunc. Praesent eget sem vel
leo ultrices bibendum. Aenean faucibus. Morbi dolor nulla, malesuada eu, pulvinar at, mollis ac, nulla. Curabitur
auctor semper nulla. Donec varius orci eget risus. Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam.
Duis eget orci sit amet orci dignissim rutrum.

\begin{eqnarray}
s(nT_{s}) &= &s(t)\times \sum\limits_{n=0}^{N-1} \delta (t-nT_{s}) \xleftrightarrow{\mathrm{DFT}}  S \left(\frac{m}{NT_{s}}\right) \nonumber\\
&= &\frac{1}{N} \sum\limits_{n=0}^{N-1} \sum\limits_{k=-N/2}^{N/2-1} s_{k} e^{\mathrm{j}2\pi k\Delta fnT_{s}} e^{-j\frac{2\pi}{N}mn}
\end{eqnarray}

\section{Sample for another first level head}\label{sec2}

Nulla malesuada porttitor diam. Donec felis erat, congue non, volutpat at, tincidunt tristique, libero. Vivamus viverra
fermentum felis. Donec nonummy pellentesque ante. Phasellus adipiscing semper elit. Proin fermentum massa ac
quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo~\cite{Rothermel1998} Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem. Sed lacinia
nulla vitae enim. Pellentesque tincidunt purus vel magna. Integer non enim. Praesent euismod nunc eu purus. Donec
bibendum quam in tellus. Nullam cursus pulvinar lectus. Donec et mi. Nam vulputate metus eu enim. Vestibulum
pellentesque felis eu massa.

Example for bibliography citations cite~\citep{Elbaum2002}, cites~\cite{Allen2011,Yoo2007}

Quisque ullamcorper placerat ipsum. Cras nibh~\cite{Yoo2007,Schulz2012} Morbi vel justo vitae lacus tincidunt ultrices. Lorem ipsum dolor sit
amet, consectetuer adipiscing elit. In hac habitasse platea dictumst. Integer tempus convallis augue. Etiam facilisis.
Nunc elementum fermentum wisi. Aenean placerat. Ut imperdiet, enim sed gravida sollicitudin, felis odio placerat
quam, ac pulvinar elit purus eget enim. Nunc vitae tortor. Proin tempus nibh sit amet nisl. Vivamus quis tortor
vitae risus porta vehicula.

Fusce mauris. Vestibulum luctus nibh at lectus. Sed bibendum, nulla a faucibus semper, leo velit ultricies tellus, ac
venenatis arcu wisi vel nisl. Vestibulum diam. Aliquam pellentesque, augue quis sagittis posuere, turpis lacus congue
quam, in hendrerit risus eros eget felis. Maecenas eget erat in sapien mattis porttitor. Vestibulum porttitor. Nulla facilisi. Sed a turpis eu lacus commodo facilisis. Morbi fringilla, wisi in dignissim interdum, justo lectus sagittis dui, et
vehicula libero dui cursus dui. Mauris tempor ligula sed lacus. Duis cursus enim ut augue. Cras ac magna. Cras nulla.
Nulla egestas. Curabitur a leo. Quisque egestas wisi eget nunc. Nam feugiat lacus vel est. Curabitur consectetuer.

\begin{figure}[t]
\centerline{\includegraphics[width=342pt,height=9pc,draft]{empty}}
\caption{This is the sample figure caption.\label{fig1}}
\end{figure}

Suspendisse vel felis. Ut lorem lorem, interdum eu, tincidunt sit amet, laoreet vitae, arcu. Aenean faucibus pede eu
ante. Praesent enim elit, rutrum at, molestie non, nonummy vel, nisl. Ut lectus eros, malesuada sit amet, fermentum
eu, sodales cursus, magna. Donec eu purus. Quisque vehicula, urna sed ultricies auctor, pede lorem egestas dui, et
convallis elit erat sed nulla. Donec luctus. Curabitur et nunc. Aliquam dolor odio, commodo pretium, ultricies non,
pharetra in, velit. Integer arcu est, nonummy in, fermentum faucibus, egestas vel, odio.

Sed commodo posuere pede. Mauris ut est. Ut quis purus. Sed ac odio. Sed vehicula hendrerit sem. Duis non
odio. Morbi ut dui. Sed accumsan risus eget odio. In hac habitasse platea dictumst. Pellentesque non elit. Fusce
sed justo eu urna porta tincidunt. Mauris felis odio, sollicitudin sed, volutpat a, ornare ac, erat. Morbi quis dolor.
Donec pellentesque, erat ac sagittis semper, nunc dui lobortis purus, quis congue purus metus ultricies tellus. Proin
et quam. Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos hymenaeos. Praesent sapien
turpis, fermentum vel, eleifend faucibus, vehicula eu, lacus.

\begin{figure*}
\centerline{\includegraphics[width=342pt,height=9pc,draft]{empty}}
\caption{This is the sample figure caption.\label{fig2}}
\end{figure*}

\subsection{Example for second level head}

Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas. Donec odio elit, dictum
in, hendrerit sit amet, egestas sed, leo. Praesent feugiat sapien aliquet odio. Integer vitae justo. Aliquam vestibulum
fringilla lorem. Sed neque lectus, consectetuer at, consectetuer sed, eleifend ac, lectus. Nulla facilisi. Pellentesque
eget lectus. Proin eu metus. Sed porttitor. In hac habitasse platea dictumst. Suspendisse eu lectus. Ut mi mi, lacinia
sit amet, placerat et, mollis vitae, dui. Sed ante tellus, tristique ut, iaculis eu, malesuada ac, dui. Mauris nibh leo,
facilisis non, adipiscing quis, ultrices a, dui.

Morbi luctus, wisi viverra faucibus pretium, nibh est placerat odio, nec commodo wisi enim eget quam. Quisque
libero justo, consectetuer a, feugiat vitae, porttitor eu, libero. Suspendisse sed mauris vitae elit sollicitudin malesuada.

Maecenas ultricies eros sit amet ante. Ut venenatis velit. Maecenas sed mi eget dui varius euismod. Phasellus aliquet
volutpat odio. Vestibulum ante ipsum primis in faucibus orci luctus et ultrices posuere cubilia Curae; Pellentesque sit
amet pede ac sem eleifend consectetuer. Nullam elementum, urna vel imperdiet sodales, elit ipsum pharetra ligula,
ac pretium ante justo a nulla. Curabitur tristique arcu eu metus. Vestibulum lectus. Proin mauris. Proin eu nunc eu
urna hendrerit faucibus. Aliquam auctor, pede consequat laoreet varius, eros tellus scelerisque quam, pellentesque
hendrerit ipsum dolor sed augue. Nulla nec lacus.

\begin{quote}
This is an example~\citep{Elbaum2002,Blanchard2015} for quote text. This is an example for quote text. This is an example for quote text. This is an example for quote text~\cite{Blanchard2015}. This is an example for quote text. This is an example for quote text. This is an example for quote text. This is an example for quote text. This is an example for quote text. This is an example for quote text~\cite{Blanchard2015}. This is an example for quote text. This is an example for quote text. This is an example for quote text.
\end{quote}

\section{Sample for next first level head}\label{sec3}

\subsection{Example for another second level head}

Suspendisse vitae elit. Aliquam arcu neque, ornare in, ullamcorper quis, commodo eu, libero. Fusce sagittis erat at
erat tristique mollis. Maecenas sapien libero, molestie et, lobortis in, sodales eget, dui. Morbi ultrices rutrum lorem.
Nam elementum ullamcorper leo. Morbi dui. Aliquam sagittis. Nunc placerat. Pellentesque tristique sodales est.
Maecenas imperdiet lacinia velit. Cras non urna. Morbi eros pede, suscipit ac, varius vel, egestas non, eros. Praesent
malesuada, diam id pretium elementum, eros sem dictum tortor, vel consectetuer odio sem sed wisi.

Sed feugiat. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Ut pellentesque
augue sed urna. Vestibulum diam eros, fringilla et, consectetuer eu, nonummy id, sapien. Nullam at lectus. In sagittis
ultrices mauris. Curabitur malesuada erat sit amet massa. Fusce blandit. Aliquam erat volutpat. Aliquam euismod.
Aenean vel lectus. Nunc imperdiet justo nec dolor.

\subsection{Second level head text}

Etiam euismod. Fusce facilisis lacinia dui. Suspendisse potenti. In mi erat, cursus id, nonummy sed, ullamcorper
eget, sapien. Praesent pretium, magna in eleifend egestas, pede pede pretium lorem, quis consectetuer tortor sapien
facilisis magna. Mauris quis magna varius nulla scelerisque imperdiet. Aliquam non quam. Aliquam porttitor quam
a lacus. Praesent vel arcu ut tortor cursus volutpat. In vitae pede quis diam bibendum placerat. Fusce elementum
convallis neque. Sed dolor orci, scelerisque ac, dapibus nec, ultricies ut, mi. Duis nec dui quis leo sagittis commodo.

\subsubsection{Third level head text}

Aliquam lectus. Vivamus leo. Quisque ornare tellus ullamcorper nulla. Mauris porttitor pharetra tortor. Sed fringilla
justo sed mauris. Mauris tellus. Sed non leo. Nullam elementum, magna in cursus sodales, augue est scelerisque
sapien, venenatis congue nulla arcu et pede. Ut suscipit enim vel sapien. Donec congue. Maecenas urna mi, suscipit
in, placerat ut, vestibulum ut, massa. Fusce ultrices nulla et nisl.

Etiam ac leo a risus tristique nonummy. Donec dignissim tincidunt nulla. Vestibulum rhoncus molestie odio. Sed
lobortis, justo et pretium lobortis, mauris turpis condimentum augue, nec ultricies nibh arcu pretium enim. Nunc
purus neque, placerat id, imperdiet sed, pellentesque nec, nisl. Vestibulum imperdiet neque non sem accumsan laoreet.
In hac habitasse platea dictumst. Etiam condimentum facilisis libero. Suspendisse in elit quis nisl aliquam dapibus.
Pellentesque auctor sapien. Sed egestas sapien nec lectus. Pellentesque vel dui vel neque bibendum viverra. Aliquam
porttitor nisl nec pede. Proin mattis libero vel turpis. Donec rutrum mauris et libero. Proin euismod porta felis.
Nam lobortis, metus quis elementum commodo, nunc lectus elementum mauris, eget vulputate ligula tellus eu neque.
Vivamus eu dolor.

Nulla in ipsum. Praesent eros nulla, congue vitae, euismod ut, commodo a, wisi. Pellentesque habitant morbi
tristique senectus et netus et malesuada fames ac turpis egestas. Aenean nonummy magna non leo. Sed felis erat,
ullamcorper in, dictum non, ultricies ut, lectus. Proin vel arcu a odio lobortis euismod. Vestibulum ante ipsum primis
in faucibus orci luctus et ultrices posuere cubilia Curae; Proin ut est. Aliquam odio. Pellentesque massa turpis, cursus
eu, euismod nec, tempor congue, nulla. Duis viverra gravida mauris. Cras tincidunt. Curabitur eros ligula, varius ut,
pulvinar in, cursus faucibus, augue.

\begin{boxtext}
\section*{Example of Boxtext}%
This is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext this is sample for boxtext.
\end{boxtext}

\paragraph{Fourth level head text}

Sed feugiat. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Ut pellentesque
augue sed urna. Vestibulum diam eros, fringilla et, consectetuer eu, nonummy id, sapien. Nullam at lectus. In sagittis
ultrices mauris. Curabitur malesuada erat sit amet massa. Fusce blandit. Aliquam erat volutpat. Aliquam euismod.
Aenean vel lectus. Nunc imperdiet justo nec dolor.

Etiam euismod. Fusce facilisis lacinia dui. Suspendisse potenti. In mi erat, cursus id, nonummy sed, ullamcorper
eget, sapien. Praesent pretium, magna in eleifend egestas, pede pede pretium lorem, quis consectetuer tortor sapien
facilisis magna. Mauris quis magna varius nulla scelerisque imperdiet. Aliquam non quam. Aliquam porttitor quam
a lacus. Praesent vel arcu ut tortor cursus volutpat. In vitae pede quis diam bibendum placerat. Fusce elementum
convallis neque. Sed dolor orci, scelerisque ac, dapibus nec, ultricies ut, mi. Duis nec dui quis leo sagittis commodo.

\subparagraph{Fifth level head text}

Aliquam lectus. Vivamus leo. Quisque ornare tellus ullamcorper nulla. Mauris porttitor pharetra
tortor. Sed fringilla justo sed mauris. Mauris tellus. Sed non leo. Nullam elementum, magna in cursus sodales, augue
est scelerisque sapien, venenatis congue nulla arcu et pede. Ut suscipit enim vel sapien. Donec congue. Maecenas
urna mi, suscipit in, placerat ut, vestibulum ut, massa. Fusce ultrices nulla et nisl.

Etiam ac leo a risus tristique nonummy. Donec dignissim tincidunt nulla. Vestibulum rhoncus molestie odio. Sed
lobortis, justo et pretium lobortis, mauris turpis condimentum augue, nec ultricies nibh arcu pretium enim. Nunc
purus neque, placerat id, imperdiet sed, pellentesque nec, nisl. Vestibulum imperdiet neque non sem accumsan laoreet.
In hac habitasse platea dictumst. Etiam condimentum facilisis libero. Suspendisse in elit quis nisl aliquam dapibus.
Pellentesque auctor sapien. Sed egestas sapien nec lectus. Pellentesque vel dui vel neque bibendum viverra. Aliquam
porttitor nisl nec pede. Proin mattis libero vel turpis. Donec rutrum mauris et libero. Proin euismod porta felis.
Nam lobortis, metus quis elementum commodo, nunc lectus elementum mauris, eget vulputate ligula tellus eu neque.
Vivamus eu dolor.

>>>>

\title{Case-Base Neural Networks: survival analysis with time-varying,
higher-order interactions}

\author[1]{Jesse Islam}

\author[2]{Maxime Turgeon}

\author[3]{Robert Sladek}

\author[4]{Sahir Bhatnagar}

\address[1]{\orgdiv{Department of Quantitative life sciences,} \orgname{McGill University}, \orgaddress{\state{Quebec}, \country{Canada}}}

\address[2]{\orgdiv{Department of Statistics}, \orgname{University of Manitoba}, \orgaddress{\state{Manitoba}, \country{Canada}}}

\address[3]{\orgdiv{Department of Human Genetics}, \orgname{McGill University}, \orgaddress{\state{Quebec}, \country{Canada}}}

\address[4]{\orgdiv{Department of Biostatistics}, \orgname{McGill University}, \orgaddress{\state{Quebec}, \country{Canada}}}

\corres{Jesse Islam's \email{jesse.islam@mail.mcgill.ca}}

@@

\abstract[Summary]{Neural network-based survival methods can model data-driven
covariate interactions. While these methods have led to better
predictive performance than regression-based approaches, they cannot
model both time-varying interactions and complex baseline hazards. To
address this, we propose Case-Base Neural Networks (CBNN) as a new
approach that combines the case-base sampling framework with flexible
architectures. Our method naturally accounts for censoring and does not
require method specific hyperparameters. Using a novel sampling scheme
and data augmentation, we incorporate time directly into a feed-forward
neural network. CBNN predicts the probability of an event occurring at a
given moment and estimates the hazard function. We compare the
performance of CBNN to survival methods based on regression and neural
networks in two simulations and two real data applications. We report
two time-dependent metrics for each model. In the simulations and real
data applications, CBNN provides a more consistent predictive
performance across time and outperforms the competing neural network
approaches. For a simple simulation with an exponential hazard model,
CBNN outperforms the other neural network methods. For a complex
simulation, which highlights the ability of CBNN to model both a complex
baseline hazard and time-varying interactions, CBNN outperforms all
competitors. The first real data application shows CBNN outperforming
all neural network competitors, while a second real data application
shows competitive performance. We highlight the benefit of combining
case-base sampling with deep learning to provide a simple and flexible
modeling framework for data-driven, time-varying interaction modeling of
survival outcomes. An R package is available at
\url{https://github.com/Jesse-Islam/cbnn}.}

\keywords{survival analysis, machine learning, case-base, neural
network}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Smooth-in-time accelerated failure time (AFT) models can estimate
absolute risks by modeling the hazard directly through a user-specified
baseline hazard distribution \citep{kleinbaum2012survival}. Cox
proportional hazards models are used more often than AFT models, causing
analyses to be based on hazard ratios and relative risks rather than on
survival curves and absolute risks \citep{hanley2009}. The
identification of an appropriate distribution for the baseline hazard in
an AFT model may be difficult for common diseases that have many
interacting risk factors, or a Cox model where the disease pathogenesis
may change with age, making it difficult to maintain the proportional
hazards assumption. For example, previous studies of breast cancer
incidence have discovered time-varying interactions with covariates of
interest, such as tumor size \citep{coradini2000time}. One approach to
provide flexibility in the baseline hazard involves using the basis of
splines on time in our model \citep{royston2002flexible}. However,
regression-based models are limited in that they require prior knowledge
about potential time-varying interactions and their quantitative
effects.

Neural networks provide a data-driven approach to approximating
interaction terms. For example, DeepSurv is a neural network-based
proportional hazards model that implements the Cox partial
log-likelihood as a custom loss function \citep{katzman2018DeepSurv},
resulting in a stepwise absolute risk curve that cannot accommodate
time-varying interactions. Compared to Cox regression, DeepSurv shows
better performance on the Study to Understand Prognoses Preferences and
Risk Treatments (SUPPORT) dataset \citep{knaus1995SUPPORT}. To handle
non-proportional hazards, a modification to the loss function was
proposed \citep{faraggi1995neural}. To remove the need for tuning the
number of layers and nodes, the concept of extreme learning machines has
been applied to a Cox neural network model \citep{wang2019extreme}.

As an alternative method that assumes a baseline hazard distribution,
DeepHit specifies each survival time of interest in the model and
directly estimates survival curves, rather than deriving a hazard
function \citep{lee2018DeepHit}. It assumes an inverse Gaussian
distribution as the baseline hazard and it outperforms DeepSurv on the
Molecular Taxonomy of Breast Cancer International Consortium (METABRIC)
dataset \citep{curtis2012genomic}.

Providing further flexibility, Deep Survival Machines (DSM) is a
parametric survival model using neural networks with a mixture of
distributions as the baseline hazard \citep{dsmPaper}. On both the
SUPPORT and METABRIC datasets, DSM outperforms DeepSurv and DeepHit
\citep{dsmPaper}. However, like DeepHit and DeepSurv, DSM cannot model
time-varying interactions.

We note that these alternative neural network approaches all require
custom loss functions \citep{katzman2018DeepSurv} \citep{lee2018DeepHit}
\citep{dsmPaper}. DeepHit introduces a hyperparameter weighing its two
loss functions (negative log-likelihood and ranking losses) while DSM
requires a two-phase learning process and user implementations for
distributions beyond Log-Normal or Weibull \citep{lee2018DeepHit}
\citep{dsmPaper}. Regression-based approaches require prior
specification of all interaction terms, which makes it challenging to
model covariate effects that change over time. The current neural
network models provide flexibility at the cost of opacity, while
regression models provide clarity at the cost of flexibility.

In this article, we propose Case-Base Neural Networks (CBNN) as a method
that models time-varying interactions and a flexible baseline hazard
using commonly available neural network components. Our approach to
modeling the full hazard uses case-base sampling \citep{hanley2009}.
This sampling technique allows probabilistic models to predict survival
outcomes. As part of the case-base framework, we use transformations of
time as a feature (covariate) to specify different baseline hazards. For
example, by including splines of time as covariates, we can approximate
the Royston-Parmar flexible baseline hazard model
\citep{royston2002flexible}, however, this still requires explicit use
of time-varying interactions. CBNN can model both without extra tuning
parameters.

In Section \ref{methods}, we describe how case-base sampling and neural
networks are combined both conceptually and algebraically, along with
our hyperparameter choices and software implementation. Section
\ref{sims}, describes our metrics and compares the performance of CBNN,
DeepSurv, DeepHit, DSM, Cox regression and case-base using logistic
regression (CBLR) on simulated data. Section \ref{casestudies} describes
the real-data analysis, while Section \ref{discussion} explores the
implications of our results and contextualizes them within neural
network survival analysis in a single event setting.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%section2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\hypertarget{methods}{%
\section{Case-base neural network methodology, metrics and
software}\label{methods}}

In this section, we define case-base sampling, which converts the total
survival time into discrete person-specific moments (person-moments).
Then, we detail how neural networks can be used within this framework,
explicitly incorporating time as a feature while adjusting for the
sampling bias. Finally, we report on the software versions used. An R
package is available for use at
\url{https://github.com/Jesse-Islam/cbnn}. The entire code base to
reproduce the figures and empirical results in this paper is available
at \url{https://github.com/Jesse-Islam/cbnnManuscript}.

\hypertarget{case-base-sampling}{%
\subsection{Case-base sampling}\label{case-base-sampling}}

Case-base sampling is an alternative framework for survival analysis
\citep{hanley2009}. In case-base sampling, we sample from the continuous
survival time of each person in our dataset to create a \emph{base
series} of \emph{person-moments}. This \emph{base series} complements
the \emph{case series}, which contains all person-moments at which the
event of interest occurs.

For each person-moment sampled, let \(X_i\) be the corresponding
covariate profile \(\left(x_{i1},x_{i2},...,x_{ip} \right)\), \(T_i\) be
the time of the person-moment and \(Y_i\) be the indicator variable for
whether the event of interest occurred at time \(T_i\). We estimate the
hazard function \(h(t \mid X_i)\) using the sampled person-moments.
Recall that \(h(t \mid X_i)\) is the instantaneous potential of
experiencing the event at time \(t\) for a given set of covariates
\(X_i\), assuming \(T_i \geq t\).

Now, let \(b\) be the (user-defined) size of the \emph{base series} and
let \(B\) be the sum of all follow-up times for the individuals in the
study. If we sample the \emph{base series} uniformly across the study
base, then the hazard function of the sampling process is equal to
\(b/B\). Therefore, we have the following equality
\footnote{We are abusing notation here, conflating hazards with probabilities. For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015} .}:
\begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align} The odds of a person-moment being a part of the \emph{case
series} is the ratio of the hazard \(h(T_i \mid X_i)\) and the uniform
rate \(b/B\). Using \eqref{eqn:main}, we can see how the log-hazard
function can be estimated from the log-odds arising from case-base
sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

To estimate the correct hazard function, we adjusting for the bias
introduced when sampling a fraction of the study base \(B\) by adding
the offset term \(\log\left(\frac{b}{B} \right)\) as in
\eqref{eqn:offset}. Next, we propose using neural networks to model the
odds.

\hypertarget{neural-networks-to-model-the-hazard-function}{%
\subsection{Neural networks to model the hazard
function}\label{neural-networks-to-model-the-hazard-function}}

After case-base sampling, we pass all features, including time, into any
user-defined feed-forward component, to which an offset term is added,
then passed through a sigmoid activation function (Figure
\ref{fig:NNarch}). We use the sigmoid function as its inverse is the
odds, which we can use to calculate the hazard. The general form for the
neural network using CBNN is:

\begin{align}\label{eqn:nnProb}
P\left(Y=1|X,T\right)&=\mathrm{sigmoid}\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) \right),
\end{align}

where \(T\) is a random variable representing the event time, \(X\) is
the random variable for a covariate profile, \(f_{\theta}(X, T)\)
represents any feed-forward neural network architecture,
\(\log\left(\frac{B}{b}\right)\) is bias term set by case-base sampling,
\(\theta\) is the set of parameters learned by the neural network and
\(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By approximating a
higher-order polynomial of time using a neural network, the baseline
hazard specification is now data-driven, where user-defined
hyperparameters such as regularization, number of layers and nodes
control the flexibility of the hazard function. We provide a detailed
description of the choices we made in the next sub-section.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../../../figures/nnarch2}}

\caption{Steps involved in CBNN from case-base sampling to the model framework we use for training. The first step is case-base sampling, completed before training begins. Next, we pass this sampled data through a feed-forward neural network. We add the offset and pass that through a sigmoid activation function, whose output is a probability. Once the neural network model completes its training, we can convert the probability output to a hazard, using it for our survival outcomes of interest.}\label{fig:NNarch}
\end{figure}

The following derivation shows how our probability estimate is converted
to odds: \begin{align*}
 \log\left( h(t \mid X) \right) &= \log\left(\frac{\mathrm{sigmoid}\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{1-\mathrm{sigmoid}\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}\right) + \log\left(\frac{b}{B}\right) \\
 &= \log\left( \frac{\frac{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)+1}}{1-\frac{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)+1}}\right) + \log\left(\frac{b}{B}\right) \\
 &= \log\left(\exp\left( f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) \right) \right) + \log\left(\frac{b}{B}\right) \\
 &= f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) + \log\left(\frac{b}{B}\right) \\
&= f_{\theta}(X, T). 
\end{align*}

We use binary cross-entropy as our loss function \citep{gulli2017}:
\begin{align*}
L(\theta)=-\frac{1}{N} \sum^{N}_{i=1} y_{i} \cdot \log(\hat{f}_{\theta}(x_{i}, t_{i}) ) + (1-y_{i} )\cdot \log(1-\hat{f}_{\theta}(x_{i}, t_{i}) ),
\end{align*} where \(\hat{f}_{\theta}(x_{i}, t_{i})\) is our estimate
for a given covariate profile and time, \(y_{i}\) is our target value
specifying whether an event occurred and \(N\) represents the number of
individuals in our training set.

Backpropagation with an appropriate minimization algorithm (e.g.~Adam,
RMSPropagation, stochastic gradient descent) is used to optimize the
parameters in the model \citep{gulli2017}. For our analysis, we use Adam
as implemented in Keras \citep{gulli2017}. Note that the size of the
\emph{case series} is fixed as the number of events, but we can make the
\emph{base series} as large as we want. A ratio of 100:1 \emph{base
series} to \emph{case series} is sufficient \citep{hanley2009}. We pass
our feed-forward neural network through a sigmoid activation function
(Figure \ref{fig:NNarch}). Finally, we can convert this model output to
a hazard. When using our model for predictions, we manually set the
offset term to 0 in the new data, as we account for the bias during the
fitting process.

Since we are directly modeling the hazard, we can readily estimate the
risk function (\(F\)) at time \(t\) for a covariate profile \(X\),
viz.\begin{align}\label{eqn:ci2}
F\left(t\mid X\right)& = 1 - \exp\left(-\int_{0}^{t}h(u|X) \,\textrm du\right).
\end{align} We use a finite Riemann sum \citep{hughes2020calculus} to
approximate the integral in \eqref{eqn:ci2}.

\hypertarget{hyperparameter-selection}{%
\subsection{Hyperparameter selection}\label{hyperparameter-selection}}

Neural networks are flexible when defining the architecture and
optimization parameters. These hyperparameter decisions can affect the
estimated parameters and were chosen during a set of initial simulations
to determine if CBNN can learn complex interactions in practice. We set
a max number of epochs to be \(2000\), batch size as \(512\), learning
rate as \(10^{-3}\), decay as \(10^{-7}\), patience as \(10\) epochs,
\(\{50,50,25,25\}\) nodes in each hidden layer with \(50\%\) dropout at
each layer, a minimum delta loss on the validation set of \(10^{-7}\)
over 10 epochs and Adam \citep{gulli2017} as our optimizer. These
choices may permit the model to approximate higher-order interactions
while preventing over-fitting \citep{srivastava2014dropout}. We fix a
train-validation-test split that allows us to update the weights with a
subset of the data (training), assess performance at each epoch
(validation) and gauge performance of the final model (test) for each
method. We select the best weights after training based on validation
loss \citep{gulli2017}.

\hypertarget{software-implementation}{%
\subsection{Software implementation}\label{software-implementation}}

R \citep{Rsoft} and python \citep{py} are used to evaluate methods from
both languages. We fit the Cox model using the \textbf{survival} package
\citep{survpkg}, the CBLR model using the \textbf{casebase} package
\citep{cbpkg}, the DeepSurv model using the \textbf{survivalmodels}
package \citep{survmods}, the DeepHit model using \textbf{pycox}
\citep{lee2018DeepHit} and the DSM model using
\textbf{DeepSurvivalMachines} \citep{dsmPaper}. We made the components
of CBNN using the \textbf{casebase} package for the sampling step and
the \textbf{keras} \citep{keras} package for our neural network
architecture. The \textbf{simsurv} package \citep{simsurv} is used for
our simulation studies, while \textbf{flexsurv} \citep{flexsurv} is used
to fit a flexible baseline hazard using splines for our complex
simulation. We use the implementation of \(C_{IPCW}\) from the python
package \textbf{sksurv} \citep{sksurv}. The \textbf{riskRegression}
package \citep{riskRegression} is used to get the Index of Prediction
Accuracy (IPA metric). Both metrics are described in detail in the
following section. We modify the \textbf{riskRegression} package to be
used with any user supplied risk function \(F\). To ensure that both R
and Python-based models are running in unison on the same data through
our simulations and bootstrap, we use the \textbf{reticulate} package
\citep{reticulate}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%section3
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\hypertarget{sims}{%
\section{Simulation studies}\label{sims}}

In this section, we use simulated data to evaluate the performance of
CBNN and compare our approach with existing regression-based (Cox, CBLR)
and neural network-based (DeepHit, DeepSurv, DSM) methods. We specify a
linear combination of each covariate as the linear predictor in
regression-based approaches (Cox, CBLR), which contrasts with neural
network approaches that allow for non-linear interactions. We simulate
data under a simple exponential model and a complex baseline hazard with
time-varying interactions, each with 10\% random censoring. For both
settings, we simulate three covariates:
